Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-10-13 main.tex
quelques remarques reviews

\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over $\Sigma$, $\Delta$, $\Semiring$ and $\bar\Phi$
%the input and output alphabets~$\Sigma$ and $\Delta$ with label theory $\bar\Phi$, and the semiring $\Semiring$
is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states,
$\mathsf{in} \colon Q \to \Semiring$   %\Phi_{\Sigma, \Delta}
(respectively $\mathsf{out} \colon Q \to \Semiring$)  %\Phi_{\Sigma, \Delta}
are functions defining the weight for entering
(respectively leaving) computation in a state,
and $\bar{\wei}$ is a triplet of transition functions
$\wei_{10}: Q \times Q \to \Phi_{\Sigma}$,
$\wei_{01}: Q \times Q \to \Phi_{\Delta}$, and
$\wei_{11}: Q \times Q \to \Phi_{\Sigma, \Delta}$.
\end{definition}
%
\noindent
%We call the \emph{number of transitions} of $T$ the number of pairs of states
%$q, q' \in Q$ such that $\wei_{10}$ or $\wei_{01}$ or $\wei_{11}$
%is not the constant $\zero$.
%
\noindent
A tuple of $Q \times (\Sigma \cup \{ \varepsilon \}) \times (\Delta \cup \{ \varepsilon \}) \times Q \to \Semiring$
is called a \emph{transition} of $T$.
For convenience, we shall sometimes present the above $\wei_{10}$, $\wei_{01}$, $\wei_{11}$
%(overloading the function names),
as mappings associating to every transition a weight value in $\Semiring$ as follows,
for every $q, q' \in Q$, $a \in \Sigma$,  $b \in \Delta$:
\[
\begin{array}{rcll}
\wei_{10}(q, a, \varepsilon, q') & = & \phi(a) & %\wei_\Sigma(q, q')(a)
\quad\mathrm{where~} \phi = \wei_{10}(q, q') \in \Phi_\Sigma,\\
\wei_{01}(q, \varepsilon, b, q') & = & \psi(b) &
\quad\mathrm{where~} \psi = \wei_{01}(q, q') \in \Phi_\Delta,\\
\wei_{11}(q, a, b, q') & = & \eta(a, b) &
\quad\mathrm{where~} \eta = \wei_{11}(q, q') \in \Phi_{\Sigma, \Delta}.\\
\end{array}
\]
%
\noindent
% The symbolic-weighted transducer
The \SWT $T$ computes on pairs $\< s, t> \in \Sigma^* \times \Delta^*$,
$s$ and $t$, being respectively called \emph{input} and \emph{output} word.
\florent{It compute asynchronously: advance in $s$ and not in $t$ ($\wei_{10}$),
or the opposite ($\wei_{01}$),
or advance in both $s$ and $t$ ($\wei_{11}$).}
% \noindent
$T$ is
based on an intermediate function $\weight_T$
defined recursively, for every states $q, q' \in Q$,
and every pairs of strings $\< s, t> \in \Sigma^* \times \Delta^*$.
% \setminus \{ \< \varepsilon, \varepsilon >\}$,
%
%$a \in \Sigma$, $u \in \Sigma^*$, $b\in \Delta$, $v\in \Delta^*$, by:
%
\begin{align}
%\weight_T(q, \varepsilon, \varepsilon, q)  & = \one \label{eq:SWT-weight}\\ %\final(q)\\
%\weight_T(q, \varepsilon, \varepsilon, q') & = \zero \quad \mathrm{if~} q \neq q'\nonumber\\
\weight_T(q, \varepsilon, \varepsilon, q')  & = \one
 \quad \mathrm{if~} q = q' \mathrm{~and~} \zero \mathrm{~otherwise} \label{eq:SWT-weight}\\
\weight_T(q, s, t, q') & = \displaystyle\bigoplus_{\begin{array}{c}
                                                   \scriptstyle q'' \in Q\\[-1ex]
                                                   \scriptstyle s = ua,\, a \in \Sigma
                                                   \end{array}}
    \weight_T(q, u, t, q'') \otimes \wei_{10}(q'', a, \varepsilon, q')\nonumber\\
                    & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                                     \scriptstyle q'' \in Q\\[-1ex]
                                                     \scriptstyle t = vb,\, b \in \Delta\\
                                                     \end{array}}
    \weight_T(q, s, v, q'') \otimes \wei_{01}(q'', \varepsilon, b, q')\nonumber\\
                    & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                                     \scriptstyle q'' \in Q\\[-1ex]
                                                     \scriptstyle s = ua,\, t = vb\\
                                                     \end{array}}
    \weight_T(q, u, v, q'') \otimes \wei_{11}(q'', a, b, q')\nonumber
\end{align}

@@

\section{SW Visibly Pushdown Automata and Best-Search}
\label{section:SWVPA}\label{sec:SWVPA}
The model presented now generalizes symbolic \VPA
(\SVPA~\cite{dAntonyAlur14SVPDA},
 themselves generalizing \VPA~\cite{AlurMadhusudan09nested} to infinite alphabets)
from Boolean semirings to arbitrary semiring domains.
It associates to every nested word over an infinite alphabet a weight value
in a semiring.
Nested words can describe structures of labeled trees.
In the context of parsing, they will be useful to
represent AST % parse trees.
(see Section~\ref{sec:parsing} and Appendix~\ref{app:trees}). % parse trees. 
%\florent{see \S 5 and App.A}

>>>>

\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over $\Sigma$, $\Delta$, $\Semiring$ and $\bar\Phi$
%the input and output alphabets~$\Sigma$ and $\Delta$ with label theory $\bar\Phi$, and the semiring $\Semiring$
is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states,
$\mathsf{in} \colon Q \to \Semiring$   %\Phi_{\Sigma, \Delta}
(respectively $\mathsf{out} \colon Q \to \Semiring$)  %\Phi_{\Sigma, \Delta}
are functions defining the weight for entering
(respectively leaving) computation in a state,
and $\bar{\wei}$ is a triplet of transition functions
$\wei_{10}: Q \times Q \to \Phi_{\Sigma}$,
$\wei_{01}: Q \times Q \to \Phi_{\Delta}$, and
$\wei_{11}: Q \times Q \to \Phi_{\Sigma, \Delta}$.
\end{definition}
%
\noindent
%We call the \emph{number of transitions} of $T$ the number of pairs of states
%$q, q' \in Q$ such that $\wei_{10}$ or $\wei_{01}$ or $\wei_{11}$
%is not the constant $\zero$.
%
\noindent
A tuple of $Q \times (\Sigma \cup \{ \varepsilon \}) \times (\Delta \cup \{ \varepsilon \}) \times Q \to \Semiring$
is called a \emph{transition} of $T$.
For convenience, we shall sometimes present the above $\wei_{10}$, $\wei_{01}$, $\wei_{11}$
%(overloading the function names),
as mappings associating to every transition a weight value in $\Semiring$ as follows,
for every $q, q' \in Q$, $a \in \Sigma$,  $b \in \Delta$:
\[
\begin{array}{rcll}
\wei_{10}(q, a, \varepsilon, q') & = & \phi(a) & %\wei_\Sigma(q, q')(a)
\quad\mathrm{where~} \phi = \wei_{10}(q, q') \in \Phi_\Sigma,\\
\wei_{01}(q, \varepsilon, b, q') & = & \psi(b) &
\quad\mathrm{where~} \psi = \wei_{01}(q, q') \in \Phi_\Delta,\\
\wei_{11}(q, a, b, q') & = & \eta(a, b) &
\quad\mathrm{where~} \eta = \wei_{11}(q, q') \in \Phi_{\Sigma, \Delta}.\\
\end{array}
\]
%
\noindent
% The symbolic-weighted transducer
The \SWT $T$ computes on pairs $\< s, t> \in \Sigma^* \times \Delta^*$,
$s$ and $t$, being respectively called \emph{input} and \emph{output} word.
\florent{It compute asynchronously: advance in $s$ and not in $t$ ($\wei_{10}$),
or the opposite ($\wei_{01}$),
or advance in both $s$ and $t$ ($\wei_{11}$).}
% \noindent
The semantics of $T$ is
based on an intermediate function $\weight_T$
defined recursively as follows, for every states $q, q' \in Q$,
and every pairs of strings $\< s, t> \in \Sigma^* \times \Delta^*$.
% \setminus \{ \< \varepsilon, \varepsilon >\}$,
%
%$a \in \Sigma$, $u \in \Sigma^*$, $b\in \Delta$, $v\in \Delta^*$, by:
%
\begin{align}
%\weight_T(q, \varepsilon, \varepsilon, q)  & = \one \label{eq:SWT-weight}\\ %\final(q)\\
%\weight_T(q, \varepsilon, \varepsilon, q') & = \zero \quad \mathrm{if~} q \neq q'\nonumber\\
\weight_T(q, \varepsilon, \varepsilon, q')  & = \one
 \quad \mathrm{if~} q = q' \mathrm{~and~} \zero \mathrm{~otherwise} \label{eq:SWT-weight}\\
\weight_T(q, s, t, q') & = \displaystyle\bigoplus_{\begin{array}{c}
                                                   \scriptstyle q'' \in Q\\[-1ex]
                                                   \scriptstyle s = ua,\, a \in \Sigma
                                                   \end{array}}
    \weight_T(q, u, t, q'') \otimes \wei_{10}(q'', a, \varepsilon, q')\nonumber\\
                    & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                                     \scriptstyle q'' \in Q\\[-1ex]
                                                     \scriptstyle t = vb,\, b \in \Delta\\
                                                     \end{array}}
    \weight_T(q, s, v, q'') \otimes \wei_{01}(q'', \varepsilon, b, q')\nonumber\\
                    & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                                     \scriptstyle q'' \in Q\\[-1ex]
                                                     \scriptstyle s = ua,\, t = vb\\
                                                     \end{array}}
    \weight_T(q, u, v, q'') \otimes \wei_{11}(q'', a, b, q')\nonumber
\end{align}

@@

\section{SW Visibly Pushdown Automata and Best-Search}
\label{section:SWVPA}\label{sec:SWVPA}
We consider now a language model generalizing symbolic \VPA
(\SVPA~\cite{dAntonyAlur14SVPDA},
 themselves generalizing \VPA~\cite{AlurMadhusudan09nested} to infinite alphabets)
from Boolean semirings to arbitrary semiring domains.
It is also a particular case of 
Weighted Symbolic Automata with Data Storage~\cite{Herrmann16dlt} for a special 
type of data storage.
It associates to every nested word over an infinite alphabet a weight value
in a semiring.
Nested words can describe structures of labeled trees.
In the context of parsing, they will be useful to
represent AST % parse trees.
(see Section~\ref{sec:parsing} and Appendix~\ref{app:trees}). % parse trees. 
%\florent{see \S 5 and App.A}
